Extended Data

### 2022-03-07 doc/manuscript/Dow_manuscript_Nature.Rmd
working on intro
#117

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*. <!-- go back and check that these references are appropriate for this statement-->
Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures, particularly for temperate deciduous forests. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@conifer_refs] and increased annual growth in mesic climates[@gao_earlier_2022;@more_conifer_refs]. 
Much less is known about the climate sensitivity of growth of deciduous broadleaf trees, although it has been observed that their growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022]. 
Furthermore, recent evidence suggests that growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
- close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016]
- start of xylogenesis makes growth possible, but actual growth depends on climate conditions during the growth period [@etzold_number_2022]
- probably include some mention of conifers (for whcih more work has been done)
- cite @gao_earlier_2022, noting that their result may be driven by conifers

@@

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs. [@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers). 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021]. <!--this sentence is repetitive with with previous.-->

@@

Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
<!--gao_earlier_2022 demonstrates the same for some other regions.-->
Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.

>>>>

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*. <!-- go back and check that these references are appropriate for this statement-->
Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs]. 
Much less is known about how deciduous broadleaf tree growth responds to spring temperatures.
While close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the timing of growth earlier alongside leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020], it has been observed that annual growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022]. 
Furthermore, recent evidence suggests that growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].

@@

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]

@@

~~While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs. [@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).~~ 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021]. <!--this sentence is repetitive with with previous.-->

@@

~~Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.~~

@@

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
<!--gao_earlier_2022 demonstrates the same for some other regions.-->
<!-- cite @gao_earlier_2022, noting that their result may be driven by conifers-->
Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.
